# Severe Weather Analysis - harmful event types

## Synopsis
This investigation explores storm data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm data base. Based on estimates of fatalities, injuries, crop and property damage we find the most harmful event types. For this analysis the relative harmfulness of an event type is being determined by comparing the total damage caused in the years included in the analysis.

Due to changes over time in the way the data has been collected only data newer than 1996 is used in the investigation.

The investigation finds that the most harmful events with respect to population health are "heat" and "tornado" with respect to both fatalities and injuries. With respect to crop and property damage the most harmful event types are "drought", "flood" and "hurricane".

## Data Processing
In this section the data base from the NOAA is prepared for further analysis. This includes downloading the database, reading in the data and filtering it to reduce the data set to the relevant measurements. Finally the data is cleaned and enriched with several new columns.

### Initialise packages
The following initialises the R packages used in the following.

```{r initialization, echo=TRUE}
library(stringr)
library(knitr)
library(ggplot2)
library(gridExtra, quietly = TRUE)
library(scales)
```

### Download data
The storm data base is downloaded from the listed URL if it does not already exists locally.

```{r download, echo=TRUE}
data.dir = "./data"
zip.filename = file.path(data.dir, "StormData.csv.bz2")
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

if(!file.exists(zip.filename)) {
    dir.create(data.dir)
    download.file(url, zip.filename, method="curl")
}

```

### Read data
The data is read from the downloaded csv.bz2 file.

```{r read, echo=TRUE, cache=TRUE}
# Note: the assignment is inconsistent on whether analysis should start from the csv file or from the csv.bz2 file. I have chosen the later.
storm.data <- read.csv(bzfile(zip.filename), 
                       colClasses=c("EVTYPE"="character", 
                                    "PROPDMGEXP"="character", 
                                    "CROPDMGEXP"="character"))
```

### Filtering
This analysis will address questions regarding population health and economic consequences. Therefore it is reasonable to only consider events where either fatalities, injuries, property damage, crop damage or any combination of these were recorded. As a result the data is filtered based on these criteria.

Furthermore, according to the [description of the data](http://www.ncdc.noaa.gov/stormevents/details.jsp "NOAA Storm Events Database description") only Tornados, Thunderstorms and Hail were reported prior to 1996. This would seem to cause an overrepresentation of these event types and skew the results. Therefore only events occuring after 1/1/1996 are included in the analysis.

```{r filter, echo=TRUE}
storm.data.filtered <- storm.data[storm.data$FATALITIES > 0 | 
                                      storm.data$INJURIES > 0 | 
                                      storm.data$PROPDMG > 0 | 
                                      storm.data$CROPDMG > 0, ]

# Add a column with a proper date 
storm.data.filtered$BEGIN.DATE <- as.Date(storm.data.filtered$BGN_DATE, format="%m/%d/%Y")

# Filter out data before 1/1/1996
storm.data.filtered <- storm.data.filtered[storm.data.filtered$BEGIN.DATE>=as.Date('1/1/1996', format="%m/%d/%Y"), ]

```

This filtering brings the number of recorded events down from **`r dim(storm.data)[1]`** to **`r dim(storm.data.filtered)[1]`**. Also it reduces the number of unique event types from **`r length(unique(storm.data$EVTYPE))`** to **`r length(unique(storm.data.filtered$EVTYPE))`**.

### Cleanup
Even with the above data reduction the number of event types is still excessive. This is largely due to casing, spelling errors, different words for the same event etc. The following section attempts to partially clean up this column. Only event types with a significant number of entries and where the change is obvious are handled.

The event type reduction is performed in the following section.

```{r cleanup, echo=TRUE}

# Clean up event type names
event.type <- storm.data.filtered$EVTYPE # Get the EVTYPE's

event.type <- tolower(event.type)
event.type <- str_trim(event.type)
event.type <- gsub('.*tstm.*|.*thunderstorm.*', 'thunderstorm', event.type)
event.type <- gsub('.*high wind.*', 'high wind', event.type)
event.type <- gsub('.*tornado.*', 'tornado', event.type)
event.type <- gsub('.*hail.*', 'hail', event.type)
event.type <- gsub('.*flash flood.*', 'flash flood', event.type)
event.type <- gsub('.*heat.*', 'heat', event.type)
event.type <- gsub('.*tropical storm.*', 'tropical storm', event.type)
event.type <- gsub('.*hurricane.*', 'hurricane', event.type)
event.type <- gsub('.*rip current.*', 'rip current', event.type)
storm.data.filtered$EVTYPE <- event.type # Set the fixed EVTYPE's back

#Just to avoid a magic number in the following
countToShow <- 10
```

The result of the above cleanup is that the data now contains **`r length(unique(storm.data.filtered$EVTYPE))`** unique event types. The `r countToShow` most recorded event types are listed in the below table: 

```{r count_types, echo=TRUE, results='asis'}
types.rle <- rle(storm.data.filtered$EVTYPE[(order(storm.data.filtered$EVTYPE))])

reported.event.counts <- data.frame(EVTYPE = types.rle$values[order(types.rle$length, decreasing = TRUE)], 
                                    COUNT = types.rle$length[order(types.rle$length, decreasing = TRUE)])

kable(head(reported.event.counts[,1:2], n=countToShow), format="markdown")
```

These `r countToShow` event types account for **`r sum(reported.event.counts[1:countToShow,2])`** events corresponding to **`r sum(reported.event.counts[1:countToShow,2])/dim(storm.data.filtered)[1]*100`%** of the filtered data. Further investigations of the event types could reduce the number of types further but it does not seem necessary for the analysis.

### Enrichment
The crop and property damages are each reported in two separate columns. E.g. the crop damage is reported in columns `CROPDMG` and `CROPDMGEXP`. One column contains the number value, the other contains the exponent. In order to easiser perform comparisons two new columns containing the converted values is added.

```{r enrichment, echo=TRUE}

# Calculating the property and crop damage from the CROPDMG, CROPDMGEXP, PROPDMG and PROPDMGEXP columns. 
# The calculation is based on mappings found on page 12 in the PDF referenced together with the data base.
# 0 -> 1
# K -> 1000
# M -> 1000000
# B -> 1000000000
exp.ref <- c("0" = 1, "K" = 1000, "M" = 1E6, "B" = 1E9)

# Set all expononts with no value to 0
clean.cropdmg.exp<-gsub('^$', 0, storm.data.filtered$CROPDMGEXP)
clean.propdmg.exp<-gsub('^$', 0, storm.data.filtered$PROPDMGEXP)

storm.data.filtered$CROP.DAMAGE <-
  unname(storm.data.filtered$CROPDMG * exp.ref[clean.cropdmg.exp])

storm.data.filtered$PROPERTY.DAMAGE <-
  unname(storm.data.filtered$PROPDMG * exp.ref[clean.propdmg.exp])


```

## Results
The following section contains the result of the data analysis. First the data is aggregated to create new data sets containing summations of the data for the four parameters used for determining population health and economic consequences.

```{r aggregate_and_sort, echo=TRUE}
fatalities.sum.per.evtype <- aggregate(FATALITIES~EVTYPE, data=storm.data.filtered, sum)
fatalities.sum.per.evtype <- fatalities.sum.per.evtype[order(fatalities.sum.per.evtype$FATALITIES, decreasing = TRUE), ]

injuries.sum.per.evtype <- aggregate(INJURIES~EVTYPE, data=storm.data.filtered, sum)
injuries.sum.per.evtype <- injuries.sum.per.evtype[order(injuries.sum.per.evtype$INJURIES, decreasing = TRUE), ]

propdmg.sum.per.evtype <- aggregate(PROPERTY.DAMAGE~EVTYPE, data=storm.data.filtered, sum)
propdmg.sum.per.evtype <- propdmg.sum.per.evtype[order(propdmg.sum.per.evtype$PROPERTY.DAMAGE, decreasing = TRUE), ]

cropdmg.sum.per.evtype <- aggregate(CROP.DAMAGE~EVTYPE, data=storm.data.filtered, sum)
cropdmg.sum.per.evtype <- cropdmg.sum.per.evtype[order(cropdmg.sum.per.evtype$CROP.DAMAGE, decreasing = TRUE), ]
```

### Population health
We can now compare the health consequences for the various event types. This is done in figure 1. 

```{r ten_most_health_consequences, echo=TRUE,fig.align='center', fig.scap='Highest health risk events since 1996'}
p1 <- ggplot(data=fatalities.sum.per.evtype[1:10,], 
    aes(x=reorder(EVTYPE, order(FATALITIES)),
        y=FATALITIES)) + 
    geom_bar(fill="#DD8888", stat="identity") +
    xlab(NULL) +
    ylab("Total fatalities") +
    coord_flip()

p2 <- ggplot(data=injuries.sum.per.evtype[1:10,], 
    aes(x=reorder(EVTYPE, order(INJURIES)),
        y=INJURIES)) + 
    geom_bar(fill="#DD8888", stat="identity") +
    xlab(NULL) +
    ylab("Total injuries") +
    coord_flip()

caption <- tableGrob(matrix(expression(
    paste("Figure 1: Event types with highest population health consequences since 1996"))),
    parse = TRUE, theme = 
        theme.list(
            gpar.corefill = gpar(fill = NA, col = NA),
            core.just = "center"
            )
    )

grid.arrange(p1,p2, sub=caption)
```

It can be seen from the figure that event types "heat" and "tornado" are the most lethal event types when examining the total number of fatalities. With respect to injuries the the event type "tornado" is clearly the event type with largest consequences having caused more than twice as many injuries as number two on the list.

### Economic consequeces
Next we compare the economic consequences for the various event types. This is done in figure 2. 

```{r economic_consequence, echo=TRUE,fig.align='center'}
p1 <- ggplot(data=cropdmg.sum.per.evtype[1:10,], 
    aes(x=reorder(EVTYPE, order(CROP.DAMAGE)),
        y=CROP.DAMAGE/1e9)) + 
    geom_bar(fill="#DD8888", stat="identity") +
    xlab(NULL) +
    ylab("Crop damage (Billion US $)") +
    coord_flip() +
    scale_y_continuous(labels = comma)

p2 <- ggplot(data=propdmg.sum.per.evtype[1:10,], 
    aes(x=reorder(EVTYPE, order(PROPERTY.DAMAGE)),
        y=PROPERTY.DAMAGE/1e9)) + 
    geom_bar(fill="#DD8888", stat="identity") +
    xlab(NULL) +
    ylab("Property damage (Billion US $)") +
    coord_flip() +
    scale_y_continuous(labels = comma)

caption <- tableGrob(matrix(expression(
    paste("Figure 2: Event types with highest economic consequences since 1996"))),
    parse = TRUE, theme = 
        theme.list(
            gpar.corefill = gpar(fill = NA, col = NA),
            core.just = "center"
            )
    )

grid.arrange(p1,p2, sub=caption)
```

The results here are more mixed than in the previous comparison. But "drought" is clearly the most damaging event type for crop while "flood" is most damaging to property. In both categories "hurricane" is the second most damaging event type.
