# Severe Weather Event Analysis - NOAA Storm Database

## Synopsis

## Data Processing

```{r initialization}
library("stringr")

```



```{r download, echo=TRUE}
data.dir = "./data"
zip.filename = file.path(data.dir, "StormData.csv.bz2")
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

if(!file.exists(zip.filename)) {
    dir.create(data.dir)
    download.file(url, zip.filename, method="curl")
}

```

```{r read, echo=TRUE, cache=TRUE}
# Note: the assignment is inconsistent on whether analysis should start from the csv file or from the csv.bz2 file. I have chosen the later.
storm.data <- read.csv(bzfile(zip.filename), 
                       colClasses=c("EVTYPE"="character"))
```

XXXX: may not really be needed. XXXThe original data file contains 985 different event types in the `EVTYPE` column. This is largely due to casing, spelling errors, different words for the same event etc. The following section attempts to clean up this column.  
```{r preprocess, echo=TRUE}

event.type <- storm.data$EVTYPE # Get the EVTYPE's

event.type <- tolower(event.type)
event.type <- str_trim(event.type)


storm.data$EVTYPE <- event.type # Set the fixed EVTYPE's back
levels(as.factor(storm.data$EVTYPE))
```

Aggregate the data to find the most severe event types.
```{r}
fatalities.sum.per.evtype <- aggregate(FATALITIES~EVTYPE, data=storm.data, sum)
fatalities.mean.per.evtype <- aggregate(FATALITIES~EVTYPE, data=storm.data, mean)

injuries.sum.per.evtype <- aggregate(INJURIES~EVTYPE, data=storm.data, sum)
injuries.mean.per.evtype <- aggregate(INJURIES~EVTYPE, data=storm.data, mean)

fatalities.sum.per.evtype <- 
    fatalities.sum.per.evtype[order(fatalities.sum.per.evtype$FATALITIES, decreasing = TRUE), ]

fatalities.mean.per.evtype <- 
    fatalities.mean.per.evtype[order(fatalities.mean.per.evtype$FATALITIES, decreasing = TRUE), ]

injuries.sum.per.evtype <- 
    injuries.sum.per.evtype[order(injuries.sum.per.evtype$INJURIES, decreasing = TRUE), ]

injuries.mean.per.evtype <- 
    injuries.mean.per.evtype[order(injuries.mean.per.evtype$INJURIES, decreasing = TRUE), ]





```




## Results